---
title: "geoprofiler"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geoprofiler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE,warning=FALSE}
library(geoprofiler)
library(ggplot2)
library(units)
library(sf)
library(dplyr)

theme_set(theme_classic())
options(ggplot2.continuous.colour = "viridis")
options(ggplot2.continuous.fill = "viridis")
```

## Load example data

For this tutorial we use the `quakes` dataset giving the locations of 1000 
seismic events of MB > 4.0. The events occurred in a cube near Fiji since 1964.
```{r load_data}
data("quakes")
crs <- st_crs("EPSG:3460")

quakes_sf <- st_as_sf(quakes, coords = c("long", "lat"), crs = "WGS84") |>
  st_transform(crs = crs)

quake_map <- ggplot() +
  geom_sf(aes(color = depth, size = mag), data = quakes_sf) +
  scale_size_binned() +
  theme_classic()

quake_map
```
> Transform your dataset into a projected cooridnate reference system to deal 
with units such as m, km, miles etc. Otherwise the units will be in degrees.


## Define a profile

There are several ways to define a profile, depending on what is known or what is more relevant.

### Profile from two known points
For example, if the profile should be a line connecting two points:
```{r pts}
profile_pts <- data.frame(lon = c(160, -170), lat = c(-15, -24)) |>
  st_as_sf(coords = c("lon", "lat"), crs = "WGS84") |>
  st_transform(crs = crs)
```

Combine the two points to a line and add the profile line to the map:

```{r line}
profile_l <- profile_line(profile_pts)

quake_map +
  geom_sf(data = profile_l, lty = 2)
```


```{r parameters}
profile_azimuth(profile_l)
profile_length(profile_l)
```


### Profile from direction and length from one point

Or, if the orientation of the profile is more relevant, we can define the profile by the direction and distance from one point: 

```{r azimuth, warning=FALSE, message=FALSE}
data.frame(lon = 160, lat = 15) |>
  st_as_sf(coords = c("lon", "lat"), crs = "WGS84") |>
  st_transform(crs = crs) |>
  profile_points(profile.azimuth = 112, profile.length = set_units(8000, km))
```
> Note that the unit of `profile.length` depends on the coordinate reference 
system and must be in degree for lon-lat cooridnates and m (km, miles, ...) if 
otherwise.




## Determine the distances along (and across) the profile

To calculate the distances along and across the profile, we simply transform the 
data into a coordinate system of the profile line:
```{r project}
quakes_profile <- profile_coords(quakes_sf, profile = profile_l) |>
  bind_cols(quakes_sf)
```

The resulting data-frame gives the distance along the profile (`X`) and the 
distance from the profile (`Y`). 

A quick way to visualize the "transformed" data can be achieved by plotting 
these axes against each other::

```{r profile_map}
quakes_profile |>
  ggplot(aes(X / 1000, Y / 1000, color = depth)) +
  geom_hline(yintercept = 0) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 3000, 250)) +
  scale_y_continuous(breaks = seq(-3000, 3000, 250)) +
  coord_fixed()
```
In this plot the profile line is a horizontal line (`X=0`). 

This plot helps also allows to easily make a subset of our data to avoid plotting data points that are too far away from the profile:

```{r filter}
quakes_profile_filtered <- filter(
  quakes_profile,
  abs(Y) <= 750 * 1000,
  X >= 1500 * 1000
)
```


## Plot data along profile
Finally we plot our filtered data against the profile:
```{r profile_plot1}
ggplot(quakes_profile_filtered, aes(X/1000, depth, color = depth, size = mag)) +
  geom_point() +
  scale_size_binned("Richter Magnitude") +
  scale_y_reverse() +
  scale_x_continuous(guide = guide_axis(position = "top")) +
  labs(x = "Distance along profile (km)", y = "Depth (km)", color = "Depth (km)")
```

One way to show the distance from the profile in the same plot is by 
controlling the size (and/or the opacity) of the points.
Here I am using the unfiltered data, and show the closest points much larger than the more distant points.
This gives a somewhat 3-dimensional look to it:

```{r profile_plot2}
ggplot(quakes_profile, aes(X/1000, depth, color = mag, size = abs(Y)/1000, alpha = abs(Y)/1000)) +
  geom_point() +
  scale_color_viridis_c("Richter Magnitude", option = "A") +
  scale_size_continuous("Distance from profile (km)", range = c(3, .1)) +
  scale_alpha_continuous("Distance from profile (km)", range = c(1, .1)) +
  scale_y_reverse() +
  scale_x_continuous(guide = guide_axis(position = "top")) +
  labs(x = "Distance along profile (km)", y = "Depth (km)")
```
